Time and space specifications for stencils

### 2016-06-19 paper.tex
expanding section on explaining the syntax

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}

@@

\newcommand{\stenFwd}[2]{\texttt{forward} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenBwd}[2]{\texttt{backward} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenCen}[2]{\texttt{centered} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenFwdS}[2]{\texttt{fwd} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenBwdS}[2]{\texttt{bwd} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}
\newcommand{\stenCenS}[2]{\texttt{cen} \, (\texttt{depth=}#1,
  \texttt{dim=}#2)}

@@

\subsection{Syntax}
\label{sec:syntax}

Figure~\ref{fig:syntax}
gives the syntax, which we introduces in stages below.
The top-level is given by the \textit{spec} production
which splits into either a \textit{spatial} specification or
a \textit{temporal specification}.
%%

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec}
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\
\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
\nonterm{spatial} ::= & [\nonterm{mod}] \;
                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive}, \; \term{dims=}\dimTy \{\term{,} \; \dimTy\} \\
\nonterm{modIrrefl} ::= & \term{irreflexive}, \; \term{dims=}\dimTy \{\term{,}
          \dimTy\}\\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy  \\
\mid \; & \term{backward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
\mid \; & \term{centered}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

\paragraph{Spatial specifications}

\paragraph{Temporal specifications}

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Basic syntax definitions} In the following, $v$
  ranges over variables, $s$ over statements, and $e$ over expressions
  (which may be impure).  By ``variable'', we mean in the imperative
  sense (\ie{}, named binders to memory cells) instead of mathematical
  variables. The function \rhsExp{} maps statements to a set of
  expressions which occur in right-hand positions (\ie{}, not the
  target of an assignment). The function \var{} maps expressions to a
  set of the variables used in right-hand positions.

\begin{definition}[Standard: base induction variable]
An integer variable \textit{i} is a \emph{base induction variable} 
within the scope of a ``for'' loop (or equivalent for the target
language, \eg{}, \texttt{do} in Fortan for our implementation) if it
is the control variable of the loop, incremented by $1$ per
iteration. 
\end{definition}

\begin{definition}[Standard: array subscript]
An \emph{array subscript} is an expression index an array, which in
our abstract we write as $v(\bar{e})$ where $\bar{e}$ is shorthand for
a syntactic list of indexing expressions. 
\end{definition}

\begin{definition}[Neighbourhood index]
An array subscript $e$ is a \emph{neighbourhood index} if,
for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
where $a$ is a constant of integer type. The relation $\equiv$ identifies terms
up-to commutativity of $+$ and the inverse
relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$).
In the following, we classify neighbourhood indices 
using the predicate \neigh{}.
\end{definition}

>>>>

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

@@

\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\subsection{Notation and convention}
\label{sec:notation}

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $v$ ranges over program variables, $s$ over statements, and
$e$ over expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables. We make clear a number of standard notions
used throughout.

\begin{definition}[Base induction variable]
  An variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop\footnote{Or equivalent for the target language, \eg{},
    \texttt{do} in Fortan for our implementation}, incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop, and such variables will be ranged
over by $i$, $j$, $k$. Since we do not use non-base induction
variables (which are traditionally affine transformations of induction
variables) we often simply say \emph{induction variable}. 
\end{definition}

\begin{definition}[Array subscript]
  An \emph{array subscript} is an expression that indexes an array
  (for the purpose or reading or writing an element), which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of some base induction variable.
\end{definition}

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form 
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\var{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a 
 \emph{neighbourhood subscript}. 
\end{definition}

\subsection{Specification syntax}
\label{sec:syntax}

Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
which we introduces in stages below.  The top-level is given by the
\nonterm{specification} production which splits into either a
\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
(specification declaration). Specification declarations associate a
specification to one or more program variables $\bar{v}$ (here given
using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$).
Specification declarations are either \nonterm{spatial} or
\nonterm{temporal} specifications and describe, for spatial
specifcations how an array variable $v$ is read, and for temporal
specifications what variables are used to define the array variable $v$.
%%

\subsubsection{Spatial specification syntax}

\emph{Regions} are the central building blocks of spatial
specifications. Regions can either be declared along with a
\nonterm{regionDec}, assigning a region specification \nonterm{region} to
a region variable \nonterm{rvar} or given directly within a
\nonterm{spatial} specification.

\paragraph{Region constants}

Regions have as terminals the \emph{region constants}, denoted by 
\term{forward}, \term{backward}, or \term{centered}. Each
region constant is given a depth parameter $n$ (natural number greater
than 0) and a dimension identifier $d$ (also a natural number greater than
0). Region constants specify that an array is read from by a collection
of indices which in the $d^{th}$ dimension are 
neighbourhood indices ranging from $i + 0$ up to $i + n$ inclusively, \eg{}, the
following is valid stencil, reading from \term{a} and writing to \term{b}:
%%
\begin{ExmVerbatim}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{ExmVerbatim}
%%
were \texttt{i} is an induction variable.  The stencil specification
is associated to the array variable \texttt{a}. Note that the second
dimension is fixed at a constant here.

A \term{backward} stencil
is similar but has negative neighbourhood indices, \eg{}, 
%
\begin{ExmVerbatim}
!= stencil backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
Note that the left-hand subscript on \term{b} does not affect
the stencil specification. The requirement on the left-hand side
is merely that it is an array subscript with at least one neighbourhood index,
but this may be on different induction variables to the right-hand
side, which is the stencil we are concerened with.

A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{ExmVerbatim}

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\
\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
\nonterm{spatial} ::= & [\nonterm{mod}] \;
                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive}, \; \term{dims=}\dimTy \{\term{,} \; \dimTy\} \\
\nonterm{modIrrefl} ::= & \term{irreflexive}, \; \term{dims=}\dimTy \{\term{,}
          \dimTy\}\\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

\paragraph{Sum and product of regions}
%%
Region terms can be combined using the sum operator
\term{+} or the product operator \term{*}.

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by both $r$ and
$r'$. It can be thought of as a conjunction of speicifcations. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) 
         + centered(depth=1, dim=2) :: a

b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) 
                   + a(i,j-1) + a(i,j+1) 
\end{ExmVerbatim}
%%
The product of two regions $r \texttt{*} r'$ specifies that the array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a nine-point stencil) has a specification as the
product of two \texttt{centered} regions in each dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) 
         * centered(depth=1, dim=2) :: a

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Note that there are a number of intermediate assignments over which 
the whole specification ranges.

\paragraph{Region declarations and variables}

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the speicifcation from the previous
 example can be restated as:
%%
\begin{ExmVerbatim}
!= region r1 = centered(depth=1, dim=1) 
!= region r2 = centered(depth=1, dim=2)
!= stencil r1*r2 :: a
\end{ExmVerbatim}
%%
\paragraph{Spatial modifiers}
%%
There are three kinds of modifier that can be given to spatial 
specifications: \texttt{readOnce}, \texttt{reflexive}, 
\texttt{irreflexive}.

The \texttt{readOnce} modifier enforces that no index appears
more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{ExmVerbatim}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
However, this would be an invalid specification if any of the 
array subscripts was reapeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts. 
The notion correponds to that of \emph{linearity}, though we
opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

The \texttt{irreflexive} modifier is parameteresed by a
comma-separated list of dimensions, indicating that for those
dimensions there is no neighbour index with $0$ offset, \eg{}, 
%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1) + a(i-1)
\end{ExmVerbatim}
%
or with just a \texttt{forward} region an example would be: 
%
\begin{ExmVerbatim}
!= region r = forward(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1)
\end{ExmVerbatim}

The \term{reflexive} modifier is dual to \term{irreflexive},
indicating dimensions in which there is a neighbour index with $0$
offset. The region constants \term{forward}, \term{backward},
\term{centered} implicitly include the \term{reflexive} modifier,
but it is useful when there is a dimension in which there is reflexive
indexing but not a region constant, \eg{},
%%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil reflexive(dims=2), r :: a
b(i,j) = a(i-1,j) + a(i+1,j) + a(i, j) 
\end{ExmVerbatim}
%%
Overlapping \term{irreflexive} and \term{reflexive} modifiers
cannot occur together-- this is an invalid specification.

\paragraph{Under and over-approximation}

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax.  For example:
%%
\begin{ExmVerbatim}
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2) 
\end{ExmVerbatim}
%%

\subsubsection{Temporal specifications}

@@

\paragraph{Syntax functions}
The function \rhsExp{} maps statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.
